Revealing transient strain in geodetic data with Gaussian process regression

### 2017-09-25 RevisedSubmission/hines_hetland_2017_gji.tex
did more

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

@@

Now that we have a posterior estimate of transient displacements, we can write the components of $\dot\varepsilon$ as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
The components of $\dot\varepsilon$ are stochastic processes, and we can consider realizations of $\dot{\varepsilon}_{ij}$ to be realizations of $\hat{u}_i$ and $\hat{u}_j$ that have undergone the operations from eq. (\ref{eq:StrainRate}). Since the operations from eq. (\ref{eq:StrainRate}) are linear, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. From \citet[sec. 10.2]{Papoulis1991}, we find that $\dot{\varepsilon}_{ij}$ has a mean function
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and covariance function
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right).
\end{equation}

>>>>

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates. We denote transient strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates, and $t$ is time. For simplicity, we assume that the study region is sufficiently small that $\vec{x}$ can be considered a point in a 2-D map projection that is aligned with the cardinal directions, $\vec{x} = (x_\mathrm{e},x_\mathrm{n})$. We determine $\dot\varepsilon$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. Transient displacements have an easting and northing component, $\vec{u} = (u_\mathrm{e}, u_\mathrm{n})$, and we make a prior assumption that each component of $\vec{u}$ is a Gaussian process with zero mean and a generic covariance function $\mathrm{cov}\left(u_i(p),u_i(p')\right) = C_{u_i}(p,p')$. Using a more concise notation, we describe each component of $\vec{u}$ as $u_i \sim \mathcal{N}\left(0,C_{u_i}\right)$. The function $C_{u_i}$ must be positive definite, which means that the matrix $[C_{u_i}(p,p')]_{(p,p') \in \mitbf{P} \times \mitbf{P}}$ is positive definite for any set of points $\mitbf{P}$ \citep[sec. 2.5]{Cressie1992}. We assume that $C_{u_i}$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u_i}\left((\vec{x},t),(\vec{x}',t')\right) = X_i(\vec{x},\vec{x}')T_i(t,t').
\end{equation}  
As long as the functions $X_i$ and $T_i$ are positive definite, $C_{u_i}$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. In order to compute transient strain rates from transient displacements, it is also necessary that the derivatives $\partial^2 X_i(\vec{x},\vec{x}')/ \partial x_j \partial x_j'$ and $\partial^2 T_i(t,t') / \partial t \partial t'$ exist everywhere. The appropriate choice for $X_i$ and $T_i$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

@@

We now have a posterior estimate of transient displacements that can be evaluated at any $p$. We can write the components of $\dot\varepsilon$ as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
The components of $\dot\varepsilon$ are stochastic processes, and we can consider realizations of $\dot{\varepsilon}_{ij}$ to be realizations of $\hat{u}_i$ and $\hat{u}_j$ that have undergone the operations from eq. (\ref{eq:StrainRate}). Since the operations from eq. (\ref{eq:StrainRate}) are linear, we know that $\dot\varepsilon_{ij}$ is also a Gaussian process. From \citet[sec. 10.2]{Papoulis1991}, we find that $\dot{\varepsilon}_{ij}$ has a mean function
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and covariance function
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right).
\end{equation}
